Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-17 manuscript_agu_rev1.tex
Merge

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model, but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction, near-surface cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

\begin{abstract}
Global storm resolving models (GSRMs) represent the next generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but necessarily utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using approximately 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500~m, associated with underestimated lower tropospheric stability and overestimated lifting condensation level. The reanalyses strongly underestimate fog and near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in MERRA-2, implying a ``too few, too bright'' cloud problem. SO cloud and fog biases are a substantial issue in the analyzed models and result in shortwave and longwave radiation biases.
\end{abstract}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, the model relies entirely on explicit simulation of cloud dynamics instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with the model and reanalyses. We subsetted our results by cyclonic activity and stability. We found that the model and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or near-surface clouds are especially underestimated in the reanalyses. However, the model and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity, and the global navigation satellite system coordinates. Derived thermodynamic (virtual potential temperature, relative humidity, lifting condensation level, etc.) and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

Some of the observatial data were likely used in reanalysis assimilation. The Macquarie Island station surface measurements and radiosonde profiles (not used in our analysis), were sent to the World Meteorological Organization Global Telecommunication System (GTS). The lidar measurements and radiosonde profiles on the RSV \textit{Aurora Australis} were not sent for assimilation.

@@

The Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2) is a reanalysis produced by the Global Modeling and Assimilation Office at the NASA Goddard Space Flight Center \cite{gelaro2017}. It uses version 5.12.4 of the Goddard Earth Observing System (GEOS) atmospheric model \cite{rienecker2008,molod2015}. Non-convective clouds (condensation, autoconversion, and evaporation) are parameterized using a prognostic scheme \cite{bacmeister2006}, and sub-grid cloud fraction is determined using total water distribution and a critical relative humidity threshold. The reanalysis output analyzed here is available at a spatial resolution of 0.5° of latitude and 0.625° of longitude, which is about 56 km in the North--South direction and 35 km in the East--West direction at 60°S. The number of vertical model levels is 72. Here, we use the following products: 1-hourly instantaneous 2D single-level diagnostics (M2I1NXASM) for 2-m temperature and humidity; 3-hourly instantaneous 3D assimilated meteorological fields (M2I3NVASM) for cloud quantities, pressure, and temperature; 1-hourly average 2D surface flux diagnostics (M2T1NXFLX) for precipitation; and 1-hourly average 2D radiation diagnostics (M2T1NXRAD) for radiation quantities \cite{merra2}. Vertically resolved fields in M2I3NVASM start at a height of about 60 m, which limits our analysis of fog and near-surface clouds in this reanalysis.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/ann.pdf}
\caption{
Artificial neural network (ANN) for prediction of precipitation in lidar backscatter. \textbf{(a)} Diagram showing the TensorFlow structure of the ANN, \textbf{(b)} randomly selected example samples of near-surface backscatter in four categories (clear, fog, rain, and snow), as determined by coincident manual weather observations, \textbf{(c)} receiver operating characteristic diagram of the ANN, \textbf{(d)} examples of 10-day time series of human-observed (``HUM'') and predicted precipitation based on an ANN trained on all voyages (``ANN'') and all voyages except for the shown voyage (``ANN2'') during three randomly selected voyages with the available data. Here, by ``randomly selected,'' we mean selected from the top of a permutation generated by a pseudo-random number generator to prevent authors' bias in the selection.
}
\label{fig:ann}
\end{figure}

@@

On the RV \emph{Polarstern} voyages, regular manual synoptic observations were available and included precipitation presence or absence and type. We used this dataset to train a convolutional artificial neural network (ANN) to recognize profiles with precipitation from lidar backscatter data (Fig.~\ref{fig:ann}a), implemented in the TensorFlow ANN framework \cite{tensorflow}. Samples of short time intervals (10~min) of near-surface lidar backscatter (0–250 m) were classified as clear, rain, snow, and fog, using the synoptic observations as a training dataset (Fig.~\ref{fig:ann}b). From these, a binary, mutually exclusive classification of profiles as precipitating (rain or snow) or dry (clear or fog) was derived. For detecting model and reanalysis precipitation, we used a fixed threshold for surface precipitation flux of 0.1 mm h$^{-1}$ (the ANN was not used).

@@

We partitioned our data into two mutually exclusive subsets by cyclonic activity. For this purpose, we used a cyclone tracking algorithm to identify extratropical and polar cyclones (ECs and PCs) over the SO in the reanalysis and ICON data. We used the open-source cyclone tracking package CyTRACK \cite{perez-alarcon2024}. Generally, what constitutes an EC is considered relatively arbitrary due to the very large variability of ECs \cite{neu2013}. The CyTRACK algorithm uses mean sea level pressure and wind speed thresholds as well as tracking across time steps to identify cyclone centers and their radii in each time step. With this information, we could classify every location at a given time as either cyclonic or non-cyclonic. Due to a relatively small total area covered by cyclones, as identified by the cyclone center and radius, for every time step and cyclone we defined a cyclonic area as a circle of double the radius identified by CyTRACK centered at the cyclone center. All other areas were defined as non-cyclonic. For identifying cyclones in the observations and the reanalyses, ERA5 pressure and wind fields were used as the input to CyTRACK. This is justified by the fact that the large-scale pressure and wind fields in ERA5 are likely sufficiently close to reality. \citeA{mcerlich2023} have shown that wind is simulated well in ERA5 relative to the WindSat polarimetric microwave radiometer measurements \cite{meissner2009}. For identifying cyclones in ICON, its own pressure and wind fields were used as the input to CyTRACK because the model is free-running, and thus the pressure and wind fields are different from reality. Subsetting by proximity to cyclones is a relatively crude measure because it does not take into account the different sectors of cyclones, which are commonly associated with different weather situations. However, this was a choice made for simplicity of the analysis, given the quantity of data. \citeA{konstali2024} performed a more complex attribution of precipitation to individual cyclone features.

@@

Fig.~\ref{fig:cyclone-stability}a, b show the geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section~\ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone and is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences, particularly in the highest occurrence rate regions, such as around Cape Adare, which is up to 20\% more cyclonic in ICON, and the Weddell and Bellingshausen Seas, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig.~\ref{fig:cyclone-stability}c, d show the geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section~\ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are common elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilizing effect \cite{knight2024}. ICON is also substantially less stable than ERA5 across the whole region. In Section~\ref{sec:thermodynamic-profiles} we show that based on radiosonde observations, the bias is in ICON and not ERA5, and it is the result of underestimated temperature at heights corresponding to 700 hPa, as well as overestimated near-surface air temperature, characterized by a higher frequency of occurrence in the 1--7°C range compared to observations at radiosonde launch locations (Fig.~\ref{fig:stats-hist-surf}a). This may be related to large-scale circulation in the model or radiative transfer biases.

@@

We used the ALCF to derive cloud occurrence by height and the total cloud fraction from observations, ICON, ERA5, and MERRA-2. The results for all campaigns individually are shown in Fig.~\ref{fig:cloud-occurrence-panel}. As shown in this figure, the biases are relatively consistent across the campaigns and longitudes. In addition, we aggregated the campaigns by calculating the averages and percentiles of all individual profiles, presented in Fig.~\ref{fig:cloud-occurrence}. The analysis shows that the total cloud fraction is underestimated in ICON by about 10\% and in the reanalyses by about 20\%. When analyzed by height, ICON overestimates cloud occurrence below 1 km and underestimates it above; MERRA-2 underestimates cloud occurrence at all heights by up to 10\%, especially near the surface; and ERA5 simulates cloud occurrence relatively well above 1 km but strongly underestimates it near the surface. We note that fog or near-surface clouds are strongly underestimated in the reanalyses (fog and clouds are both included in the cloud occurrence). We conclude that the ICON results match the observations better than the reanalyses in this metric.

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10--30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the lifting condensation level (LCL) distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface relative humidity often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air--sea moisture flux parameterization, or weaker stability in the models can cause more boundary mixing across heights and thus lower near-surface RH.

When the data are subset by latitude (Fig.~\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone (40--55°S) displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone (between 55°S and the Antarctic coast), and this could be because higher latitudes have greater prevalence of weakly stable profiles (Fig.~\ref{fig:cyclone-stability}c, d), although more stable profiles populate regions south of 65°S close to the Antarctic coast. Cyclonic activity is also stronger in high-latitude SO, which is typically associated with shallow or deep convection rather than very stable stratification necessary for fog formation. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (near-surface cloud occurrence is still strongly underestimated).

@@

In all subsets, even when the models overestimate cloud occurrence at some altitudes, they always substantially underestimate the total cloud fraction. ICON can be generally characterized as substantially overestimating cloud occurrence below 1~km and underestimating above, underestimating the total cloud fraction, and showing the greatest biases in conditions of weak stability and non-cyclonic conditions. ICON also has a peak cloud occurrence at higher altitudes than observations (500~m vs. near the surface), and correspondingly, its LCL tends to be higher. MERRA-2 can be generally characterized as underestimating cloud occurrence at nearly all altitudes as well as the total cloud fraction, but mostly above and below 500~m (the peak at 500~m is well represented). MERRA-2 displays the largest errors relative to observations in the low-latitude SO zone and under weak stability. ERA5 can be generally characterized as representing cloud occurrence correctly above about 1.5~km, overestimating between 500~m and 1~km, but underestimating near-surface cloud occurrence (0--500~m). The total cloud fraction is strongly underestimated in all subsets. ERA5 has a tendency towards greater cloud underestimation in the low-latitude SO zone and under weak stability; conversely, it overestimates the peak of cloud occurrence at 500~m in the high-latitude SO zone and under strong stability.

@@

We note that the radiative transfer calculations used in the lidar simulator mean that the impact of both cloud phase and cloud fraction are convolved to produce the cloud mask. Therefore, the cloud occurrence is not affected by any cloud phase biases as long as the cloud is optically thick enough to be detected and the laser signal is not too attenuated. An overestimated cloud albedo (in cloudy areas) can be due to either cloud liquid and ice water content, cloud phase, droplet or ice crystal size distribution, shape or orientation of ice crystals, cloud overlap, or their combination. The influence of cold clouds is likely second-order due to the much larger typical effective radius of ice crystals than cloud droplets.

@@

In all the subsets (Fig.~\ref{fig:cloud-occurrence}b--g), the same type of biases are observed, namely the outgoing SW radiation is underestimated in ICON and ERA5 and overestimated in MERRA-2, and the outgoing LW radiation is overestimated in all the models. The total cloud fraction is higher by 6\% over the high-latitude SO than the low-latitude SO, and the outgoing SW radiation is much greater by 41~Wm$^{-2}$. ICON has little difference in the total cloud fraction between low- and high-latitude SO, but greater outgoing SW radiation by 14~Wm$^{-2}$ over the high-latitude SO, likely due to thicker clouds and deeper convection in less stable and more cyclonic conditions relative to the low-latitude SO. In contrast, the reanalyses showed both greater total cloud fraction and outgoing SW radiation over the high-latitude SO compared to the low-latitude SO.

>>>>

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model, but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction, very low-level ($<$~250~m) cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

\begin{abstract}
Global storm resolving models (GSRMs) represent the next generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but necessarily utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using approximately 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500~m, associated with underestimated lower tropospheric stability and overestimated lifting condensation level. The reanalyses strongly underestimate fog and very low-level clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in MERRA-2, implying a ``too few, too bright'' cloud problem. SO cloud and fog biases are a substantial issue in the analyzed models and result in shortwave and longwave radiation biases.
\end{abstract}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subsetted our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the models (ICON, MERRA-2, and ERA5), we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. Derived thermodynamic [virtual potential temperature ($theta_v$), RH, lifting condensation level, etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

Some of the observatial data were likely used in reanalysis assimilation. The Macquarie Island station surface measurements and radiosonde profiles (not used in our analysis) were sent to the World Meteorological Organization Global Telecommunication System (GTS). The surface-level, lidar and radiosonde measurements on the RSV \textit{Aurora Australis} were not used outside of research purposes.

@@

The Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2) is a reanalysis produced by the Global Modeling and Assimilation Office at the NASA Goddard Space Flight Center \cite{gelaro2017}. It uses version 5.12.4 of the Goddard Earth Observing System (GEOS) atmospheric model \cite{rienecker2008,molod2015}. Non-convective clouds (condensation, autoconversion, and evaporation) are parameterized using a prognostic scheme \cite{bacmeister2006}, and sub-grid cloud fraction is determined using total water distribution and a critical RH threshold. The reanalysis output analyzed here is available at a spatial resolution of 0.5° of latitude and 0.625° of longitude, which is about 56 km in the North--South direction and 35 km in the East--West direction at 60°S. The number of vertical model levels is 72. Here, we use the following products: 1-hourly instantaneous 2D single-level diagnostics (M2I1NXASM) for 2-m temperature and humidity; 3-hourly instantaneous 3D assimilated meteorological fields (M2I3NVASM) for cloud quantities, pressure, and temperature; 1-hourly average 2D surface flux diagnostics (M2T1NXFLX) for precipitation; and 1-hourly average 2D radiation diagnostics (M2T1NXRAD) for radiation quantities \cite{merra2}. Vertically resolved fields in M2I3NVASM start at a height of about 60 m, which limits our analysis of fog and very low-level ($<$~250~m) clouds in this reanalysis.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/ann.pdf}
\caption{
Artificial neural network (ANN) for prediction of precipitation in lidar backscatter. \textbf{(a)} Diagram showing the TensorFlow structure of the ANN, \textbf{(b)} randomly selected example samples of very low-level (0–250~m) backscatter in four categories (clear, fog, rain, and snow), as determined by coincident manual weather observations, \textbf{(c)} receiver operating characteristic diagram of the ANN, \textbf{(d)} examples of 10-day time series of human-observed (``HUM'') and predicted precipitation based on an ANN trained on all voyages (``ANN'') and all voyages except for the shown voyage (``ANN2'') during three randomly selected voyages with the available data. Here, by ``randomly selected,'' we mean selected from the top of a permutation generated by a pseudo-random number generator to prevent authors' bias in the selection.
}
\label{fig:ann}
\end{figure}

@@

On the RV \emph{Polarstern} voyages, regular manual synoptic observations were available and included precipitation presence or absence and type. We used this dataset to train a convolutional artificial neural network (ANN) to recognize profiles with precipitation from lidar backscatter data (Fig.~\ref{fig:ann}a), implemented in the TensorFlow ANN framework \cite{tensorflow}. Samples of short time intervals (10~min) of very low-level lidar backscatter (0–250 m) were classified as clear, rain, snow, and fog, using the synoptic observations as a training dataset (Fig.~\ref{fig:ann}b). From these, a binary, mutually exclusive classification of profiles as precipitating (rain or snow) or dry (clear or fog) was derived. For detecting model and reanalysis precipitation, we used a fixed threshold for surface precipitation flux of 0.1 mm h$^{-1}$ (the ANN was not used).

@@

We partitioned our data into two mutually exclusive subsets by cyclonic activity. For this purpose, we used a cyclone tracking algorithm to identify extratropical and polar cyclones (ECs and PCs) over the SO in the reanalysis and ICON data. We used the open-source cyclone tracking package CyTRACK \cite{perez-alarcon2024}. Generally, what constitutes an EC is considered relatively arbitrary due to the very large variability of ECs \cite{neu2013}. The CyTRACK algorithm uses mean sea level pressure and wind speed thresholds as well as tracking across time steps to identify cyclone centers and their radii in each time step. With this information, we could classify every location at a given time as either cyclonic or non-cyclonic. Due to a relatively small total area covered by cyclones, as identified by the cyclone center and radius, for every time step and cyclone we defined a cyclonic area as a circle of double the radius identified by CyTRACK centered at the cyclone center. All other areas were defined as non-cyclonic. For identifying cyclones in the observations and the reanalyses, ERA5 pressure and wind fields were used as the input to CyTRACK. This is justified by the fact that the large-scale pressure and wind fields in ERA5 are likely sufficiently close to reality. \citeA{mcerlich2023} have shown that wind is simulated well in ERA5 relative to the WindSat polarimetric microwave radiometer measurements \cite{meissner2009}. For identifying cyclones in ICON, its own pressure and wind fields were used as the input to CyTRACK because ICON is free-running, and thus the pressure and wind fields are different from reality. Subsetting by proximity to cyclones is a relatively crude measure because it does not take into account the different sectors of cyclones, which are commonly associated with different weather situations. However, this was a choice made for simplicity of the analysis, given the quantity of data. \citeA{konstali2024} performed a more complex attribution of precipitation to individual cyclone features.

@@

Fig.~\ref{fig:cyclone-stability}a, b show the geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section~\ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone and is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and ICON agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences, particularly in the highest occurrence rate regions, such as around Cape Adare, which is up to 20\% more cyclonic in ICON, and the Weddell and Bellingshausen Seas, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that ICON is free-running.

Fig.~\ref{fig:cyclone-stability}c, d show the geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section~\ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are common elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilizing effect \cite{knight2024}. ICON is also substantially less stable than ERA5 across the whole region. In Section~\ref{sec:thermodynamic-profiles} we show that based on radiosonde observations, the bias is in ICON and not ERA5, and it is the result of underestimated temperature at heights corresponding to 700 hPa, as well as overestimated near-surface (2~m) air temperature, characterized by a higher frequency of occurrence in the 1--7°C range compared to observations at radiosonde launch locations (Fig.~\ref{fig:stats-hist-surf}a). This may be related to large-scale circulation in ICON or radiative transfer biases.

@@

We used the ALCF to derive cloud occurrence by height and the total cloud fraction from observations, ICON, ERA5, and MERRA-2. The results for all campaigns individually are shown in Fig.~\ref{fig:cloud-occurrence-panel}. As shown in this figure, the biases are relatively consistent across the campaigns and longitudes. In addition, we aggregated the campaigns by calculating the averages and percentiles of all individual profiles, presented in Fig.~\ref{fig:cloud-occurrence}. The analysis shows that the total cloud fraction is underestimated in ICON by about 10\% and in the reanalyses by about 20\%. When analyzed by height, ICON overestimates cloud occurrence below 1 km and underestimates it above; MERRA-2 underestimates cloud occurrence at all heights by up to 10\%, especially near the surface; and ERA5 simulates cloud occurrence relatively well above 1 km but strongly underestimates it near the surface. We note that fog or very low-level clouds are strongly underestimated in the reanalyses (fog and clouds are both included in the cloud occurrence). We conclude that the ICON results match the observations better than the reanalyses in this metric.

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10--30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the lifting condensation level (LCL) distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air--sea moisture flux parameterization, or weaker stability in the models can cause more boundary mixing across heights and thus lower near-surface RH.

When the data are subset by latitude (Fig.~\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone (40--55°S) displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone (between 55°S and the Antarctic coast), and this could be because higher latitudes have greater prevalence of weakly stable profiles (Fig.~\ref{fig:cyclone-stability}c, d), although more stable profiles populate regions south of 65°S close to the Antarctic coast. Cyclonic activity is also stronger in high-latitude SO, which is typically associated with shallow or deep convection rather than very stable stratification necessary for fog formation. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (very low-level cloud occurrence is still strongly underestimated).

@@

In all subsets, even when the models overestimate cloud occurrence at some altitudes, they always substantially underestimate the total cloud fraction. ICON can be generally characterized as substantially overestimating cloud occurrence below 1~km and underestimating above, underestimating the total cloud fraction, and showing the greatest biases in conditions of weak stability and non-cyclonic conditions. ICON also has a peak cloud occurrence at higher altitudes than observations (500~m vs. near the surface), and correspondingly, its LCL tends to be higher. MERRA-2 can be generally characterized as underestimating cloud occurrence at nearly all altitudes as well as the total cloud fraction, but mostly above and below 500~m (the peak at 500~m is well represented). MERRA-2 displays the largest errors relative to observations in the low-latitude SO zone and under weak stability. ERA5 can be generally characterized as representing cloud occurrence correctly above about 1.5~km, overestimating between 500~m and 1~km, but underestimating very low-level cloud occurrence. The total cloud fraction is strongly underestimated in all subsets. ERA5 has a tendency towards greater cloud underestimation in the low-latitude SO zone and under weak stability; conversely, it overestimates the peak of cloud occurrence at 500~m in the high-latitude SO zone and under strong stability.

@@

We note that the radiative transfer calculations used in the lidar simulator mean that the impact of both cloud phase and cloud fraction are convolved to produce the cloud mask. Therefore, the cloud occurrence is not affected by any cloud phase biases as long as the cloud is optically thick enough to be detected and the laser signal is not too attenuated. A combination of underestimated total cloud fraction and overestimated outgoing SW at TOA is indicative of an overestimated cloud albedo (in cloudy areas) due to either cloud liquid and ice water content, cloud phase, droplet or ice crystal size distribution, shape or orientation of ice crystals, cloud overlap, or their combination. The influence of cold clouds is likely second-order due to the much larger typical effective radius of ice crystals than cloud droplets.

@@

In all the subsets (Fig.~\ref{fig:cloud-occurrence}b--g), the same type of biases are observed, namely the outgoing SW radiation is underestimated in ICON and ERA5 and overestimated in MERRA-2, and the outgoing LW radiation is overestimated in all the models. Even though the total cloud fraction is higher by 6\% over the high-latitude SO than the low-latitude SO, the outgoing SW radiation is much greater by 41~Wm$^{-2}$, implying a much greater cloud albedo (of cloudy areas) over the high-latitude SO. ICON has little difference in the total cloud fraction between low- and high-latitude SO, but greater outgoing SW radiation by 14~Wm$^{-2}$ over the high-latitude SO, likely due to thicker clouds under deeper convection in less stable and more cyclonic conditions relative to the low-latitude SO. In contrast, the reanalyses showed both greater total cloud fraction and outgoing SW radiation over the high-latitude SO compared to the low-latitude SO.
